---
title: "H-1B Visa Application"
bibliography: reference.bib
format: 
  html:
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "Contents"
    embed-resources: true
number-sections: true
execute:
  echo: false
  warning: false
  message: false
knitr:
  opts_chunk:
    fig-align: center
    fig-cap: true
    fig-cap-location: bottom 
editor: visual
---

```{r}
#| label: import-packages

library(tidyverse)
library(scales)
library(knitr)
library(here)
```

# Introduction

The H-1B is a non -immigrant visa program in the United States that allows employers to employ foreign workers. The program is aimed towards professionals with a "theoretical and practical applications of a body of highly specialized knowledge" This program allows companies in the United States to hire foreign professionals with specific skills and expertise across various industries and locations [@uscis_h1b_employer_data_hub].

```{r}
#| label: import-data

visa <- read.csv2("Employer Information.csv",
                  header = TRUE,
                  sep = "\t",
                  fileEncoding = "UTF-16LE",
                  encoding = "UTF-8")
visa <- visa[, -1]
```

# Analysis

## Trends

The analysis of H-1B applications for the fiscal years, 2020 - 2024, reveals some patterns:

```{r}
#| label: fig-approvalRate
#| fig-cap: "The total number of H-1B Visa Approvals"

yearlyCounts <- visa %>%
  group_by(Fiscal.Year) %>%
  summarise(approval = sum(Initial.Approval == 1),
            denial = sum(Initial.Denial),
            .groups = "drop")
ggplot(yearlyCounts, aes(x = Fiscal.Year, y = approval)) +
  geom_col() + 
  labs(title = "H-1B Approval",
       x = "Year",
       y = "Approval Amount") + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5),
        panel.background = element_rect(fill = "white"),
        panel.grid.minor = element_line(color = "grey"))
```

```{r}
#| label: fig-denialRate
#| fig-cap: "The total number of H-1B Visa Denials"

ggplot(yearlyCounts, aes(x = Fiscal.Year, y = denial)) + 
  geom_col() + 
  labs(title = "H-1B Denial",
       x = "Year",
       y = "Denial Amount") + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5),
        panel.background = element_rect(fill = "white"),
        panel.grid.minor = element_line(color = "grey"))
```

```{r}
#| label: tbl-summaryStats
#| tbl-cap: "Summary Statistics of H-1B Visa Program"

yearlyCounts <- yearlyCounts %>%
  mutate(total_applications = approval + denial,
         approvalRate = (approval / total_applications) * 100,
         denialRate = (denial / total_applications) * 100)

kable(view(round(yearlyCounts)))
```

```{r}
#| label: tbl-testing
#| tbl-cap: "Chisqaure Test"
table <- matrix(c(yearlyCounts$approval, 
                  yearlyCounts$denial),
                ncol = 2,
                byrow = FALSE)
colnames(table) <- c("Approval", "Denial")
rownames(table) <- yearlyCounts$Fiscal.Year

test <- chisq.test(table)
test
```

Overall, the approval rates have increased while denials have decreased. The trend suggest a shift in H-1B visa processing policies and practices over the 5 year period, with a clear move towards higher approval rates and fewer denials.

The chi-square test helps to determine if the observed changes in approval and denial rates are statically significant or merely random chance. This is crucial because:

-   it validates the trends
-   provides evidence to support the claims about the changes

The p-value, \<2.2e-16, suggests that the approval and denial rate is not due too random chance.

## Distribution

### Distribution per State (Top 5 State)

```{r}
#| label: fig-perState
#| fig-cap: "Distribution per State"

stateDist <- visa %>%
  group_by(Petitioner.State) %>%
  summarise(totalApplication = n()) %>%
  arrange(desc(totalApplication)) %>%
  head(5)
  

ggplot(stateDist, aes(x = reorder(Petitioner.State, -totalApplication), y = totalApplication)) + 
  geom_col() + 
  labs(title = "Top 5 States for H-1B Applications",
       x = "State",
       y = "Total Applications") + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5),
        panel.background = element_rect(fill = "white"),
        panel.grid.minor = element_line(color = "grey"))
```

### Distribution per Industry (Top 5 Industries)

```{r}
#| label: fig-perIndustry
#| fig-cap: "Distribution per Industry"
industryDist <- visa %>%
  group_by(Industry..NAICS..Code) %>%
  summarise(totalApplication = n()) %>%
  arrange(desc(totalApplication)) %>%
  head(5)

ggplot(industryDist, aes(x = reorder(Industry..NAICS..Code, -totalApplication), y = totalApplication)) + 
  geom_col() + 
  coord_flip() + 
  labs(title = "Top 5 Industy's for H-1B Applications",
       x = "Industry",
       y = "Total Application") + 
  theme_minimal() + 
  theme(plot.title = element_text(hjust = 0.5),
        panel.background = element_rect(fill = "white"),
        panel.grid.minor = element_line(color = "grey"))
```

# Conclusion

The approval rates have been increasing over the last 5 years with the denials decreasing, indicating that more and more foreign workers are getting access to the United States. The denial rate during the 2020 fiscal year could be due to the COVID-19 pandemic, which caused major disruptions in the visa processing.

The statistical evidence strengthens the conclusion that there has been a substantial shift in the H-1B processing, with a clear trend towards higher approval rates and lower denial rates.
